---
title: "Take-home Exercise 1"
author: "Brigitta Karen Tsai"
date: "February 8, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
---

# Ship Performance Analysis

# 1 Overview

## 1.1 The Background

## 1.2 The Data

The following are the details of the data source for this analysis:

1.  Data Source: [Ship Performance Clustering Dataset](https://www.kaggle.com/datasets/jeleeladekunlefijabi/ship-performance-clustering-dataset)
2.  Dataset Size: 18 columns x 2,736 rows
3.  Data format: csv file

## 1.3 The Task

As the graphical editor of the media company, an article about ship performance in the Gulf of Guinea will be written and published. Hence, this paper will include the following contents:

1.  Exploratory Data Analysis of ship performance based on different underlying factors
2.  Visualization of key operational metrics of the ship for future performance optimizations in the maritime industry

# 2 Getting Started

## 2.1 Load and Install Packages

For this analysis, **tidyverse** and **ggplot2** packages will be used.

```{r}
pacman::p_load(tidyverse, ggrepel, patchwork, ggthemes, hrbrthemes, ggiraph,
               plotly, DT, readxl, gifski, gapminder, gganimate, ggdist, ggridges,
               colorspace, ggstatsplot, crosstalk, FunnelPlotR, knitr, scales,
               corrplot)
```

## 2.2 Import Data

Import ship performance dataset to get started with the analysis.

```{r}
ship <- read_csv("data/Ship.csv")
```

# 3 Data Preparation

## 3.1 Select Variables

Before we begin with the analysis, let us do some variable analysis and check the columns that we will be using for this analysis. Each row of the dataset represents the **performance metrics** and **attributes** for a specific **voyage/ship** over a given timeframe.

```{r}
colnames(ship)
```

For easier read, let us list down all the columns in the table below to analyse the data types and the description.

::: tabset-pane
### Ship Dataframe

::: {.table .table-striped .table-hover .table-bordered}
| Column Name             | Type           | Description                              |
|---------------------|-------------------|--------------------------------|
| Date                    | date           | timestamp of data entry                  |
| Ship_Type               | categorical    | types of vessel                          |
| Route_Type              | categorical    | types of route                           |
| Engine_Type             | categorical    | types of ship engine                     |
| Maintenance_Status      | categorical    | current maintenance status               |
| Speed_Over_Ground_knots | numerical      | average speed (knots)                    |
| Engine_Power_kW         | numerical      | engine power output (kW)                 |
| Distance_Traveled_nm    | numerical      | total distance travelled (nm)            |
| Draft_meters            | numerical      | draft of the vessel (m)                  |
| Weather_Condition       | categorical    | weather condition during operations      |
| Cargo_Weight_tons       | numerical      | weight of goods on the ship (tons)       |
| Operational_Cost_USD    | numerical      | total operational cost per voyage (USD)  |
| Revenue_per_Voyage_USD  | numerical      | total revenue generated per voyage (USD) |
| Turnaround_Time_hours   | numerical      | time taken per voyage (hours)            |
| Efficiency_nm_per_kWh   | numerical      | energy efficiency (nm/kWh)               |
| Seasonal_Impact_Score   | numerical      | *not listed in Kaggle*                   |
| Weekly_Voyage_Count     | integer        | total no. of voyage per week             |
| Average_Load_Percentage | percentage (%) | *not listed in Kaggle*                   |
:::
:::

## 3.2 Remove Columns

From the table, we will keep all the categorical and numerical columns to gain further insights on this dataset. There are 3 columns that we should remove from this dataset, which are **Date, Seasonal_Impact_Score** and **Average_Load_Percentage.**

1.  **Date**: This column is removed because it only consists of the timestamp of the data entry, which is not necessarily value-add to this analysis.
2.  **Seasonal_Impact_Score**: This column is removed because there is no description of how the value in this column is calculated. As there is no details on the column calculation, there will be no insights gained from analysing this column.
3.  **Average_Load_Percentage**: This column is removed because there is no description and source of the column calculation, hence, there is no need to use this column for the analysis.

The 3 columns will be removed using the code chunk below:

```{r}
ship <- ship %>% select(-c("Date", "Seasonal_Impact_Score", "Average_Load_Percentage"))
```

```{r}
ship
```

## 3.3 Handling Missing Value

### 3.3.1 Categorical Column

```{r}
table(ship$Ship_Type)
```

```{r}
table(ship$Route_Type)
```

```{r}
table(ship$Engine_Type)
```

```{r}
table(ship$Maintenance_Status)
```

```{r}
table(ship$Weather_Condition)
```

```{r}
ship <- ship[ship$Ship_Type != "None", ]
ship <- ship[ship$Route_Type != "None", ]
ship <- ship[ship$Engine_Type != "None", ]
ship <- ship[ship$Maintenance_Status != "None", ]
ship <- ship[ship$Weather_Condition != "None", ]
```

```{r}
dim(ship)
```

### 3.3.2 Numerical Column

In this section, we will further clean numerical data type column to ensure the entire dataset is clean. The code chunks below will check the sum of null values of each numerical column.

```{r}
sum(is.na(ship$Speed_Over_Ground_knots))
```

```{r}
sum(is.na(ship$Engine_Power_kW))
```

```{r}
sum(is.na(ship$Distance_Traveled_nm))
```

```{r}
sum(is.na(ship$Draft_meters))
```

```{r}
sum(is.na(ship$Cargo_Weight_tons))
```

```{r}
sum(is.na(ship$Operational_Cost_USD))
```

```{r}
sum(is.na(ship$Revenue_per_Voyage_USD))
```

```{r}
sum(is.na(ship$Turnaround_Time_hours))
```

```{r}
sum(is.na(ship$Efficiency_nm_per_kWh))
```

```{r}
sum(is.na(ship$Weekly_Voyage_Count))
```

```{r}
glimpse(ship)
```

average operational cost per ship type, average revenue per ship type, stats of voyage per week

## 3.4 Check for Outliers in Numerical Column

There are total of 10 numerical columns that we can use for our statistical analysis. In this section, we will do an individual column check to see whether there is outliers affecting the column of the dataset.

```{r}
boxplot.stats(ship$Speed_Over_Ground_knots)$out
```

```{r}
boxplot.stats(ship$Engine_Power_kW)$out
```

```{r}
boxplot.stats(ship$Distance_Traveled_nm)$out
```

```{r}
boxplot.stats(ship$Draft_meters)$out
```

```{r}
boxplot.stats(ship$Operational_Cost_USD)$out
```

```{r}
boxplot.stats(ship$Cargo_Weight_tons)$out
```

```{r}
boxplot.stats(ship$Revenue_per_Voyage_USD)$out
```

```{r}
boxplot.stats(ship$Turnaround_Time_hours)$out
```

```{r}
boxplot.stats(ship$Efficiency_nm_per_kWh)$out
```

```{r}
boxplot.stats(ship$Weekly_Voyage_Count)$out
```

The result from the codes above shows that there are no potential outliers affecting the dataset.

## 3.5 Derive New Columns

To gain more insights on this analysis, let us derive a new column to find out the profit gained on each voyage of the ships. The code chunk below will use **mutate()** to derive the new column.

```{r}
ship <- ship %>% 
  mutate(Profit_per_Voyage_USD = Revenue_per_Voyage_USD - Operational_Cost_USD)
```

# 4 Exploratory Data Analysis

## 4.1 Profit Analysis

```{r}
profit <- ship[ship$Profit_per_Voyage_USD >= 0, ]
```

Let us plot a the distribution of profit earned per voyage using the histogram below.

::: panel-tabset
### Plot

```{r, echo=FALSE}
v1 <- ggplot(data = ship, aes(x = Profit_per_Voyage_USD)) + 
  geom_histogram(bins = 50,
                 color = "black",
                 fill = "grey") + 
  coord_cartesian(xlim=c(-500000,1000000),
                  ylim=c(0,80)) + 
  scale_x_continuous("Profit (USD)",               
                     breaks = seq(-500000, 1000000, by = 250000),
                     labels = scales::comma) +
  labs(
    title = "Profit Distribution of Ships per Voyage (USD)",
    y = "Count"
  ) + 
  theme_minimal()

ggplotly(v1)
```

The histogram shows a normal distribution ranging from USD -500,000 to USD 1,000,000. The distribution of profit is symmetrical with the highest count at USD 250,000. Normal distribution shows that it is uncommon to have a very high profit, and it is also uncommon to get extreme losses from ship journey. The x-axis shows there are presence of negative profit value, ranging from USD 0 to USD -500,000. Although the occurrence is not as often, is is still a questionable event of why certain ships incur losses in the ship operations. The histogram roughly highlights financial variability in the ship and maritime industry.

### Code

```{r, eval=FALSE}
v1 <- ggplot(data = ship, aes(x = Profit_per_Voyage_USD)) + 
  geom_histogram(bins = 50,
                 color = "black",
                 fill = "grey") + 
  coord_cartesian(xlim=c(-500000,1000000),
                  ylim=c(0,80)) + 
  scale_x_continuous("Profit (USD)",               
                     breaks = seq(-500000, 1000000, by = 250000),
                     labels = scales::comma) +
  labs(
    title = "Profit Distribution of Ships per Voyage (USD)",
    y = "Count"
  ) + 
  theme_minimal()

ggplotly(v1)
```
:::

## 4.2 Loss Analysis

### 4.2.1 Loss Incurred by Ship Types

After the distribution plot, we are interested to find out the reason of loss incurred in the maritime industry. Let us create a new dataframe to do further analysis on the losses.

```{r}
loss <- ship[ship$Profit_per_Voyage_USD < 0, ]
```

::: panel-tabset
### Plot

```{r, echo=FALSE}
tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean Loss:", mean, "+/-", sem)
}

v2 <- ggplot(data=loss, 
                   aes(x = Ship_Type),
) +
  stat_summary(aes(y = Profit_per_Voyage_USD, 
                   tooltip = after_stat(  
                     tooltip(y, ymax))),  
    fun.data = "mean_se", 
    geom = GeomInteractiveCol,  
    fill = "light blue"
  ) +
  stat_summary(aes(y = Profit_per_Voyage_USD),
    fun.data = mean_se,
    geom = "errorbar", width = 0.2, size = 0.2
  ) + 
  labs(
    title = "Loss per Voyage by Ship Type",
    x = "Ship Type",
    y = "Profit/Loss per Voyage (USD)"
  )

girafe(ggobj = v2,
       width_svg = 8,
       height_svg = 8*0.618)
```

The plot suggests that bulk carrier incurs the most losses as compared to other types of ship. Bulk carrier has a mean loss of **USD 181,054** with a margin of error of USD 10,185. Other ship types incur a mean loss ranging from USD 140,000 to 160,000. From this plot, we are curious on why bulk carriers incur such a big loss in operating their ship.

### Code

```{r, eval=FALSE}
tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean Loss:", mean, "+/-", sem)
}

v2 <- ggplot(data=loss, 
                   aes(x = Ship_Type),
) +
  stat_summary(aes(y = Profit_per_Voyage_USD, 
                   tooltip = after_stat(  
                     tooltip(y, ymax))),  
    fun.data = "mean_se", 
    geom = GeomInteractiveCol,  
    fill = "light blue"
  ) +
  stat_summary(aes(y = Profit_per_Voyage_USD),
    fun.data = mean_se,
    geom = "errorbar", width = 0.2, size = 0.2
  ) + 
  labs(
    title = "Loss per Voyage by Ship Type",
    x = "Ship Type",
    y = "Profit/Loss per Voyage (USD)"
  )

girafe(ggobj = v2,
       width_svg = 8,
       height_svg = 8*0.618)
```
:::

### 4.2.2 Bulk Carrier Analysis

From the previous plot, we want to further deep dive into bulk carrier ship to find out the reason of such a big loss in operating this ship. Hence, let us derive a new dataframe by filtering out only the bulk carriers data using the code below.

```{r}
bc <- ship[ship$Ship_Type == "Bulk Carrier",]
```

::: panel-tabset
### Plot

```{r, echo=FALSE}
#| fig-width: 12
#| fig-height: 8
v3 <- ggplot(bc, 
       aes(x = Engine_Type, 
           y = Operational_Cost_USD)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA,
               position = position_nudge(x = -0.2)) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  scale_y_continuous("Operational Cost (USD)",               
                     breaks = seq(0, 500000, by = 100000),
                     labels = scales::comma) +
  labs(
    title = "Operational Cost Distribution of Bulk Carrier by Engine Type (USD)"
  ) + 
  coord_flip() +
  theme_minimal()

v3
```

There are 3 types of engines that could be used to operate a bulk carrier, **Steam Turbine**, **Heavy Fuel Oil** and **Diesel**. The boxplot suggests that, on average, it requires more cost to use Heavy Fuel Oil. The half eye graph shows the distribution shape of the operational cost. Heavy Fuel Oil and Steam Turbine have higher peaks, while Diesel has lower density distribution. This means that the cost of using Diesel is more spread out as compared to other engine types.

### Code

```{r, echo=FALSE}
#| fig-width: 12
#| fig-height: 8
v3 <- ggplot(bc, 
       aes(x = Engine_Type, 
           y = Operational_Cost_USD)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA,
               position = position_nudge(x = -0.2)) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  scale_y_continuous("Operational Cost (USD)",               
                     breaks = seq(0, 500000, by = 100000),
                     labels = scales::comma) +
  labs(
    title = "Operational Cost Distribution of Bulk Carrier by Engine Type (USD)"
  ) + 
  coord_flip() +
  theme_minimal()

v3
```
:::

## 4.3 Ship Type and Route Type Analysis

### 4.3.1 Cargo Weight of Each Ship Type

### 4.3.2 Loss Incurred by Each Ship Type and Route Type

In this section, we will further breakdown each ship type according to the route type and analyse the profit and loss incurred from each voyage.

::: panel-tabset
### Plot

```{r, echo=FALSE}
v4 <- ggplot(data = loss, 
            aes(x = Ship_Type, fill = Route_Type)) +
  geom_bar(stat = "count", 
           position = "dodge", 
           width = 0.7) +
  scale_y_continuous(limits = c(0, 50)) +
  scale_fill_manual(values = c("steelblue", "slategray", "darkslategray", "lightsteelblue")) +
  theme_minimal() +
  labs(x = "Ship Type", 
       y = "Count", 
       title = "Ship Type Distribution by Route Type that Incurred Loss") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggplotly(v4)
```

This code visualizes the count of ship type by each route type that incurred losses during the ship operations. The chart suggests that container ship that uses long-haul route incurred the most losses as compared to other types of route. The second most loss incurred is bulk carrier that uses coastal route and also short-haul. Container ships and bulk carrier typically weighs significantly higher than other types of ship. Hence, the high operational cost that outweighs the revenue of the ship operations.

### Code

```{r, eval=FALSE}
v4 <- ggplot(data = loss, 
            aes(x = Ship_Type, fill = Route_Type)) +
  geom_bar(stat = "count", 
           position = "dodge", 
           width = 0.7) +
  scale_y_continuous(limits = c(0, 50)) +
  scale_fill_manual(values = c("steelblue", "slategray", "darkslategray", "lightsteelblue")) +
  theme_minimal() +
  labs(x = "Ship Type", 
       y = "Count", 
       title = "Ship Type Distribution by Route Type that Incurred Loss") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggplotly(v4)
```
:::

```{r}
v7 <- ggplot(data = ship, 
            aes(x = Weather_Condition)) +
  geom_bar_interactive(
    aes(tooltip = paste0("Count: ", after_stat(count))), 
    width = 0.7, 
    fill = "steelblue"
  ) +
  facet_wrap(~ Route_Type, scales = "free") +
  scale_y_continuous(limits = c(0, 200)) +
  theme_minimal() +
  labs(x = "Weather Condition", 
       y = "Count", 
       title = "Weather Condition Distribution by Ship Type") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip()

girafe(
  ggobj = v5,
  width_svg = 6,
  height_svg = 6*0.618
)
```

## 4.4 Route Type against Cost

```{r}

```
