---
title: "Take-home Exercise 1"
author: "Brigitta Karen Tsai"
date: "February 8, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
---

# Ship Performance Analysis

# 1 Overview

## 1.1 Background

## 1.2 The Data

The following are the details of the data source for this analysis:

1.  Data Source: [Ship Performance Clustering Dataset](https://www.kaggle.com/datasets/jeleeladekunlefijabi/ship-performance-clustering-dataset)
2.  Dataset Size: 18 columns x 2,736 rows
3.  Data format: csv file

## 1.3 The Task

As the graphical editor of the media company, an article about ship performance in the Gulf of Guinea will be written and published. Hence, this paper will include the following contents:

1.  Exploratory Data Analysis of ship performance based on different underlying factors
2.  Visualization of key operational metrics of the ship for future performance optimizations in the maritime industry

# 2 Getting Started

## 2.1 Load and Install Packages

For this analysis, **tidyverse** and **ggplot2** packages will be used.

```{r}
pacman::p_load(tidyverse, ggplot2)
```

## 2.2 Import Data

Import ship performance dataset to get started with the analysis.

```{r}
ship <- read_csv("data/Ship.csv")
```
